Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2018-02-11 article3c.Rmd
introduce marginal modeling

---
title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    number_sections: yes
    toc: no
    toc_depth: 2
    keep_tex: true
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
titlepage: yes
bibliography: "mybib.bib"
documentclass: article
classoption: titlepage
fontsize: 12pt
citation_package: natbib
biblio-style: "apalike"
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{graphicx}
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class. The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality. The results show that high social class is not always favourable for survival. Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern. We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
---

@@

Another practical problem is the size of the data set:
It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
 `r n.d + pn.d` deaths. This makes it highly inconvenient to use the currently so popular semi-parametric technique known as \emph{Cox regression} [@cox72]. Instead we use a parametric model
that allows for a reduction of the data to \emph{sufficient statistics} [@fisher22]. The idea is to keep all relevant covariates as \emph{categorical} and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three marital statuses, ten periods, two
urban/rural groups, four social classes,
ten age groups, and 22 cohorts; in all, 105,600 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 14,084. The data set is thus reduced in
size by 99 per cent.

@@

@@

>>>>

---
title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    number_sections: yes
    toc: no
    toc_depth: 2
    keep_tex: true
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
titlepage: yes
bibliography: "mybib.bib"
documentclass: article
classoption: titlepage
fontsize: 12pt
citation_package: natbib
biblio-style: "apalike"
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{graphicx}
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class. The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality. The results show that high social class is not always favourable for survival. Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern. We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology. In this version of article3, we are adopting a marginal approach, in contrast to earlier versions, where the approach was conditional."
---

@@

Another practical problem is the size of the data set:
It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
 `r n.d + pn.d` deaths. This makes it highly inconvenient to use the currently so popular semi-parametric technique known as \emph{Cox regression} [@cox72]. Instead we use a parametric model
that allows for a reduction of the data to \emph{sufficient statistics} [@fisher22]. The idea is to keep all relevant covariates as \emph{categorical} and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have 2 sexes, 3 marital statuses, 12 periods, 2
urban/rural groups, 5 social classes,
14 age groups, and 22 cohorts; in all, 221,760 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 17,181. The data set is thus reduced in
size by 99 per cent.

## The conditional model

@@

## The marginal model

A alternative, and preferable, approach is the marginal model, where the disturbace effect of confounders is 
balanced out through weighting. In our case , `civil status` and `urban/rural` are confounders, that is, they 
are correlated to both our outcome, mortality, and our explanatory variable, social class. The latter association 
manifests itself through the facts that the marriage frequency varies social class and certain social classes are 
more common in the rural area than others (we find farmers on the countryside).

@@

### Association between civil status and social class over time

In order to simplify the illustration, we look at `civil status` in two groups, `ever married` and `unmarried`, in Figure \@ref(fig:assesciv).

```{r assesciv, fig.cap = "Fraction ever married by hisclass and period.", fig.height = 4}
orr <- aggregate(x = vb0Tab$exposure, by = vb0Tab[, c("period", "civst", "hisclass")], FUN = sum)

levper <- levels(orr$period)
nper <- length(levper)
levhis <- levels(orr$hisclass)
nhis <- length(levhis)
out <- matrix(0, nrow = nhis, ncol = nper)
for (per in 1:nper){
    for (his in 1:nhis){
        oo <- orr[orr$period == levper[per] & orr$hisclass == levhis[his], ]
        nam <- sum(oo$x)
        out[his, per] <- sum(oo$x[oo$civst != "unmarried"]) / nam
    }
}
out <- 100 * out
xli <-  c(1812, 1837, 1862, 1887, 1912, 1932, 1978, 1982, 1988, 1995, 2005, 2011)
plot(xli, out[1, ], type = "b", xlab = "", ylab = "Fraction ever married", 
     axes = FALSE, col = scol[1], lty = 1, pch = 1, ylim = c(0,100))
axis(1, at = xli, labels = levper, las = 2, cex = 0.7)
axis(2, las = 1)
box()
abline(h = 0)
for (r in 2:nhis){
    lines(xli, out[r, ], col = scol[r], lty = r, pch = r, type = "b")
}
legend("bottomright", legend = levhis, lty = 1:5, col = scol, pch = 1:5)
```

### Association between urban/rural and social class over time

```{r assesurb, fig.cap = "Fraction urban by hisclass and period.", fig.height = 4}
orr <- aggregate(x = vb0Tab$exposure, by = vb0Tab[, c("period", "urban", "hisclass")], FUN = sum)

levper <- levels(orr$period)
nper <- length(levper)
levhis <- levels(orr$hisclass)
nhis <- length(levhis)
out <- matrix(0, nrow = nhis, ncol = nper)
for (per in 1:nper){
    for (his in 1:nhis){
        oo <- orr[orr$period == levper[per] & orr$hisclass == levhis[his], ]
        nam <- sum(oo$x)
        out[his, per] <- sum(oo$x[oo$urban == TRUE]) / nam
    }
}
out <- 100 * out
xli <-  c(1812, 1837, 1862, 1887, 1912, 1932, 1978, 1982, 1988, 1995, 2005, 2011)
plot(xli, out[1, ], type = "b", xlab = "", ylab = "Fraction urban", 
     axes = FALSE, col = scol[1], lty = 1, pch = 1, ylim = c(0,100))
axis(1, at = xli, labels = levper, las = 2, cex = 0.7)
axis(2, las = 1)
box()
abline(h = 0)
for (r in 2:nhis){
    lines(xli, out[r, ], col = scol[r], lty = r, pch = r, type = "b")
}
legend("bottomleft", legend = levhis, lty = 1:5, col = scol, pch = 1:5)
```

### Inverse frequency weighting

In order to eliminate the disturbing effect on the association between hisclass and mortality by
the confounders civil status and urban/rural, weights that eliminates the associations between hisclass 
and the confounders are introduced. This is done period by period.

As an illustration, take the period `1901-1925` with males. There are six combinations of civil status and urban/rural, and for 
each we calculate the total exposure time. The total distribution by confounders is

```{r exweight}
exm <- vb0Tab[vb0Tab$period == "1901-1925" & vb0Tab$sex == "male", c("age", "civst", "urban", "exposure", "event", "hisclass")]
all <- aggregate(exm["exposure"], by = exm[c("civst", "urban")], FUN = sum)
all
```

and for the elite we have

```{r forelite}
elite <- exm[exm$hisclass == "elite", c("age", "civst", "urban", "exposure", "event", "hisclass")]
el <- aggregate(elite["exposure"], by = elite[, c("civst", "urban")], FUN = sum)
el
```

The weights are now 'all$exposure / el$exposure`, that is

```{r wei}
all$wei <- all$exposure / el$exposure
all
```

Put these weights on the data frame `elite`:

```{r putonel}
all$id <- paste(all$civst, all$urban, sep = "_")
elite$id <- paste(elite$civst, elite$urban, sep = "_")
indx <- match(elite$id, all$id)
elite$wght <- all$wei[indx]
```

Now repeat this for the other four hisclasses:

```{r therest}
middle <- exm[exm$hisclass == "middle", c("age", "civst", "urban", "exposure", "event", "hisclass")]
el <- aggregate(middle["exposure"], by = middle[, c("civst", "urban")], FUN = sum)
all$wei <- all$exposure / el$exposure
middle$id <- paste(middle$civst, middle$urban, sep = "_")
indx <- match(middle$id, all$id)
middle$wght <- all$wei[indx]
el
##
worker <- exm[exm$hisclass == "worker", c("age", "civst", "urban", "exposure", "event", "hisclass")]
el <- aggregate(worker["exposure"], by = worker[, c("civst", "urban")], FUN = sum)
all$wei <- all$exposure / el$exposure
worker$id <- paste(worker$civst, worker$urban, sep = "_")
indx <- match(worker$id, all$id)
worker$wght <- all$wei[indx]
el
##
farmer <- exm[exm$hisclass == "farmer", c("age", "civst", "urban", "exposure", "event", "hisclass")]
el <- aggregate(farmer["exposure"], by = farmer[, c("civst", "urban")], FUN = sum)
all$wei <- all$exposure / el$exposure
farmer$id <- paste(farmer$civst, farmer$urban, sep = "_")
indx <- match(farmer$id, all$id)
farmer$wght <- all$wei[indx]
el
##
non <- exm[exm$hisclass == "NA", c("age", "civst", "urban", "exposure", "event", "hisclass")]
el <- aggregate(non["exposure"], by = non[, c("civst", "urban")], FUN = sum)
all$wei <- all$exposure / el$exposure
non$id <- paste(non$civst, non$urban, sep = "_")
indx <- match(non$id, all$id)
non$wght <- all$wei[indx]
el
```

And put together:

```{r puttog}
per1901 <- rbind(elite, middle, worker, farmer, non)
per1901$age <- factor(per1901$age)
```

Let us fit with and without weights. Without weights:

```{r testfit, fig.cap = "Survival by hisclass, men 1901-1925, non-weighted analysis.", fig.height = 4}
fit1 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "elite", ], family = poisson)
co1 <- fit1$coefficients
co1[2:length(co1)] <- co1[2:length(co1)] + co1[1]
co1 <- exp(co1)
##
fit2 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "middle", ], family = "poisson")
co2 <- fit2$coefficients
co2[2:length(co2)] <- co2[2:length(co2)] + co2[1]
co2 <- exp(co2)
##
fit3 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "worker", ], family = "poisson")
co3 <- fit3$coefficients
co3[2:length(co3)] <- co3[2:length(co3)] + co3[1]
co3 <- exp(co3)
##
fit4 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "farmer", ], family = "poisson")
co4 <- fit4$coefficients
co4[2:length(co4)] <- co4[2:length(co4)] + co4[1]
co4 <- exp(co4)
##
fit5 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "NA", ], family = "poisson")
co5 <- fit5$coefficients
co5[2:length(co5)] <- co5[2:length(co5)] + co5[1]
co5 <- exp(co5)
###
cuts <- seq(5, 55, by = 5)
x <- seq(0, 60, length = 1000)
s1 <- ppch(x, cuts = cuts, levels = co1[1:12], lower.tail = FALSE)
plot(x, s1, type = "l", col = scol[1], axes = FALSE, xlab = "age", ylab = "Surviving fraction after 30.")
axis(1, at = seq(0, 60, by = 10), labels = seq(30, 90, by = 10))
box()
axis(2, las = 1)
abline(h = 0)
abline(v = 0)
s2 <- ppch(x, cuts = cuts, levels = co2[1:12], lower.tail = FALSE)
lines(x, s2, col = scol[2], lty = 2)
s3 <- ppch(x, cuts = cuts, levels = co3[1:12], lower.tail = FALSE)
lines(x, s3, col = scol[3], lty = 3)
s4 <- ppch(x, cuts = cuts, levels = co4[1:12], lower.tail = FALSE)
lines(x, s4, col = scol[4], lty = 4)
s5 <- ppch(x, cuts = cuts, levels = co5[1:12], lower.tail = FALSE)
lines(x, s5, col = scol[5], lty = 5)
##
```

With weights, see Figure \@ref(fig:testfit2):

```{r testfit2, fig.cap = "Survival by hisclass, men 1901-1925, weighted analysis.", fig.height = 4}
fit1 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "elite", ], family = poisson, weights = wght)
co1 <- fit1$coefficients
co1[2:length(co1)] <- co1[2:length(co1)] + co1[1]
co1 <- exp(co1)
##
fit2 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "middle", ], family = "poisson", weights = wght)
co2 <- fit2$coefficients
co2[2:length(co2)] <- co2[2:length(co2)] + co2[1]
co2 <- exp(co2)
##
fit3 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "worker", ], family = "poisson", weights = wght)
co3 <- fit3$coefficients
co3[2:length(co3)] <- co3[2:length(co3)] + co3[1]
co3 <- exp(co3)
##
fit4 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "farmer", ], family = "poisson", weights = wght)
co4 <- fit4$coefficients
co4[2:length(co4)] <- co4[2:length(co4)] + co4[1]
co4 <- exp(co4)
##
fit5 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "NA", ], family = "poisson", weights = wght)
co5 <- fit5$coefficients
co5[2:length(co5)] <- co5[2:length(co5)] + co5[1]
co5 <- exp(co5)
###
cuts <- seq(5, 55, by = 5)
x <- seq(0, 60, length = 1000)
s1 <- ppch(x, cuts = cuts, levels = co1[1:12], lower.tail = FALSE)
plot(x, s1, type = "l", col = scol[1], axes = FALSE, xlab = "Age", ylab = "Surviving fraction after 30")
axis(1, at = seq(0, 60, by = 10), labels = seq(30, 90, by = 10))
axis(2, las = 1)
box()
abline(h = 0)
abline(v = 0)
s2 <- ppch(x, cuts = cuts, levels = co2[1:12], lower.tail = FALSE)
lines(x, s2, col = scol[2], lty = 2)
s3 <- ppch(x, cuts = cuts, levels = co3[1:12], lower.tail = FALSE)
lines(x, s3, col = scol[3], lty = 3)
s4 <- ppch(x, cuts = cuts, levels = co4[1:12], lower.tail = FALSE)
lines(x, s4, col = scol[4], lty = 4)
s5 <- ppch(x, cuts = cuts, levels = co5[1:12], lower.tail = FALSE)
lines(x, s5, col = scol[5], lty = 5)
##
```
